Rice yield improvements by plant breeding are offset by decreases over time

### 2017-06-21 output/breed_paper.Rnw
Commit before messing with figures

\documentclass[12pt,]{elsarticle}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile, textcomp}
\usepackage{setspace}
\usepackage{etoolbox}
\doublespacing
\usepackage{lineno}
\usepackage{lscape} %% Harvard
\bibliographystyle{model2-names}\biboptions{authoryear}
\pagenumbering{gobble}
%% For figure numbers in supplementary info
%% from http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
\newcommand{\beginsupplement}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
}
\usepackage{Sweave}
\SweaveOpts{prefix.string=graphics/report}

@@

\title{Rice yield improvements by plant breeding are offset by
  decreases over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\author[ucd]{David MacKill}
\author[ucd]{Jim E. Hill}
\author[res]{Kent McKenzie}
\author[ucd]{Bruce A. Linquist}

\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}

@@

  Using the variety trial data, we find consistent estimates of the
  reduction of yield once a variety is released, estimated to be between
  -12 to -22 kg per year. This effect persists using all data, only
  named/released cultivars, or only released cultivars only at the Rice
  experiment station. This effect is offset by yield improvements over
  the years, estimated to be small but positive effect of

@@

\begin{keyword} rice, yield potential, temperature
\end{keyword}

@@

\linenumbers

@@

load("../data/model_data.rda")
load("h1_fit.Rda")
load("../data/all_vt_weather.Rda")

RES = res = mod_data$site == "RES"

@@

release_yr_coef = round(fixef(RES_full)[2] * scl, 1)
yrs_coef = round(fixef(RES_full)[3] * scl, 1)

rel_CI = round(posterior_interval(RES_full, pars = "release_yr_c") * scl, 1)
m202_v_206 = round(release_yr_coef * 19, 0)

@

@@

Constraints on ariable land are increasing simultaneous with the need
to increase total production to meet growing demand \citep{Foley2011a,
Godfray2010, Mueller2012}, which has necessitated harvesting more
grain per unit land area \citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the Mathesian disaster through continued
yield improvement of staple crops, however more and more production
systems are experiencing plateaus in grain yield \citep{Grassini2013,
Peng2010}. If continued improvements per unit land area are no longer
possible, increasing the area under cultivation will be needed to meet
demand, which carries with it undesirable ecological
implications. Therefore, it is of critical importance to better
understand why grain yields have not continued to increase in
instensified production systems.

California (CA) rice production represents one such rice production
system. Rice yields in CA during the period from 1920 -- 1990
increased rapidly, in part due to improvements in rice genetics (i.e.,
the Green Revolution) through the efforts of the California
Cooperative Rice Research Foundation (CCRRF; a collaboration between
the University of California, the USDA-Agricultural Research Service,
and farmer-funded research). However, since 1990 CA state average
yields have increased more slowly (Fig. \ref{fig:state_yield}) despite
demonstrated improvements of grain yield as the result of plant
breeding \citep{ccrrf}, with each new cultivar having a yield
advantage over the previous cultivars in side-by-side comparisons in
the annual Statewide Variety Trials. Yet, taken across years, these
same reports show stagnant yields; for example, at the year of release
(1985), cultivar M-202 averaged 11 Mg ha\textsuperscript{-1}
\citep[14\% MC;][]{m202}. In 2015, the average yield of the newest
released cultivar M-209 was 10.8 Mg ha\textsuperscript{-1}
\citep{m209}. In other systems, it has been shown that rice cultivars
experience a lack of adaptation to changing biotic and abiotic
conditions, resulting in a decrease of a cultivar's performance over
time \citep{Peng2010, Peng2000, Datta1995}. We sought to explore this
phenomonon and better understand how grain yields have increased in
the breeders' fields without a corresponding increase in yield in the
farmers' fields. Two competing hypotheses were tested: (1) the yield
advantage offered by newly developed cultivars is not greater than
previous cultivars, or (2) yield increases gained through plant
breeding are being offset by decreases in the yield of released
cultivars over time.

\begin{figure}[ht]
<<fig=TRUE, echo=FALSE>>=
CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))/1000
CA_data = CA_data[order(CA_data$Year),]
plot(Value * 1.12 ~ Year, CA_data,
     # type = "l",
     pch = 16,
     xlab = "Year",
     ylab = expression("Average grain yield ( Mg ha"^-1~")"))
mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
                  CA_data$Value * 1.12,
                  rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
                rep(1/11,11))
lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)

@
\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
  -- \Sexpr{max(CA_data$Year)} for California, USA. Rapid gains in
    grain yield were experienced from approximately 1950 -- 1990, but
    annual gains since 1990 have been slower. The line represents the
    10 year moving average.}
\label{fig:state_yield}
\end{figure}

@@

The Rice Experiment Station (RES) of the CCRRF is located north-west of
Biggs, CA (39.4648, -121.7342), and has been the central location for
CCRRF's efforts to develop improved rice varieties adapted for CA since
1969. The climate at the RES is Mediterranian, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} -- \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% SOM \citep{ssurgo} in the
top 15cm. As the central location for the breeding program, the RES is
researcher-managed (as opposed to farmer-managed trials) and conducts
the bulk of the trials for the Statewide Variety Trials (approximately
3 -- 4 times the number of plots compared to any other location). For
these reasons, we focused our investigation on the RES to avoid
potential complications due to site, climate, and management
differences at the other 6 -- 8 farmer-managed trials per year.

@@

Plots at the RES are managed similar to perscribed best management
practice, including land preparation, fertility management, and pest
and disease control. For planting, rice seed is pre-germinated prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\textsuperscript{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
moisture content prior to reporting.

@@

Yield was modeled as a response to the number of years since the
variety was officially released using a mixed-effects linear
model. Differences between years and cultivars were included in the

%
% model as random effects according to:

@@

Data was processed and models were fit in R, an environment for
statistical computing \citep{R}. Models were fit using the 'rstanarm'
package \citep{rstanarm}, an interface to Stan, a language for
probabilistic programming \citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results.

@@

@@

\begin{figure}[ht]

@@

plot_data = prep_plot_data(mod_data$yield_kg[res] / 1000,
                           mod_data$yrs_in_trial[res])

@@

@
\caption{The relationship between yield and the years since release
  for medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. The estimated decrease over time once a
  cultivar is release is estimated to be
  \Sexpr{yrs_coef} kg ha\textsuperscript{-1}
  year\textsuperscript{-1}. The shaded area is the 90\% credible
  interval.}
\label{fig:RES_year}
\end{figure}

\begin{figure}[ht]
<<fig=TRUE, echo=FALSE>>=
yrs <- 1975:2025
pp <- posterior_linpred(RES_full, re.form = NA,
                        newdata = data.frame(release_yr_c = yrs - 2000,
                                             yrs_in_trial = 0),
                        prob = 0.5)
pp <-  (pp * scl + center) / 1000
ints = apply(pp, 2, quantile, p = c(0.05, 0.95))

res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_kg[res]/1000 ,
                           mod_data$id[res])

plot_data$tmp = tapply(mod_data$release_yr[res],
                       mod_data$id[res], unique)

plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = expression("Grain yield ( Mg ha"^-1~")"),
     ylim = c(8, 12.5),
     xlim = c(1980,2019))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * release_yr_coef)) / 1000,
       b = release_yr_coef/1000)
abline(lm(yield_kg/1000 ~ release_yr, data = mod_data[res,]), lty = 2)
with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr, y = plot_data$yield_kg + 0.15,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.8)

legend("bottomright", legend = c("w/ years since release adj.",
                                 "w/o years since release adj.",
                                 "90% Credible interval"),
       lty = c(1,2, NA),
       bty = "n",
       fill = c(NA,NA, "grey"),
       border = NA,
       merge = TRUE,
       cex = 0.8)
polygon(x = c(yrs, rev(yrs)),
        y = c(ints[1,],rev(ints[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))

# text(1990, 8.0, bquote(.(round(release_yr_coef))~kg~ha^-1~year^-1))

@

\caption{The yield increase over time of medium grain rice varieties
  in California, USA. Ignoring the years since release (dashed line)
  suggests an increase in yield potential over time. However, correcting
  the comparison to account for differences in the length of time since
  each cultivar was released (solid line) shows no increase in yield over time.}
\label{fig:release}
\end{figure}

\begin{figure}[ht]
<<fig=TRUE, width = 8, height = 5, out.width = '85%', echo=FALSE>>=
tars <- c("M202","M206")
par(mfrow = c(1, length(tars)),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

@@

@
\caption{The performance of two major cultivars at the CA Rice
  Experiment Station (RES) over the period 1995 -- 2015. The solid line
  represents a simple linear regression of yield for each cultivar over
  the period which it was planted at the RES (including only that
  cultivar's data), while the dashed line is the average yield of M-202
  at its year of release (1985). Both cultivars have a similar yield at
  their respective year of release, and similar decreases over
  time. The slope for the individual regressions are comparible to the
slope estimated from all medium grain cultivars
(\Sexpr{yrs_coef} kg ha\textsuperscript{-1} year\textsuperscript{-1}).}
\label{fig:compare_yrs}
\end{figure}

<<echo=FALSE>>=
RES_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
@

When yields are compared accounting for differences in the years since
release at the time of comparision, there is a small decrease in yield
over time once a cultivar is released
(\Sexpr{yrs_coef}
kg ha\textsuperscript{-1} year\textsuperscript{-1}, 90\%
CI: \Sexpr{RES_CI[2]} -- \Sexpr{RES_CI[1]}). The oldest cultivar in
the trials is estimated to have lost \Sexpr{round(yrs_coef*36,1)}
kg ha\textsuperscript{-1} since release. Again
comparing the yield of two prominant cultivars, M-202 and M-206, the
two cultivars are experiencing a similar trend over time to eachother
(Fig. \ref{fig:compare_yrs}) and the estimate from all released medium
grains. The estimated yield advantage of the newer cultivar (M-206)
(\Sexpr{m202_v_206} kg ha\textsuperscript{-1}) can be explained by the
differences in the amount of time that has passed since each cultivar
was released (Fig. \ref{fig:compare_yrs} \&
\ref{fig:compare_since_release})

\subsection{Yield improvements over time}

<<echo=FALSE>>=
# m202_v_206 = round((posterior_interval(RES_full, pars = "release_yr_c") * scl) * 19, 0)

# m202_v_206 = ((posterior_predict(RES_full, newdata = data.frame(release_yr_c = c(-15, 4)),
#                                  re.form = NA) * scl) + center)
# m202_v_206 = m202_v_206[,2] - m202_v_206[,1]
# m202_v_206 = round(quantile(m202_v_206, p = c(0.05, 0.5, 0.95)), 1)

@

Contrary to the estimates for decreases once a cultivar is released,
there was no increase in yield potential over time once the
differences between the years since release were accounted for
(Fig. \ref{fig:release}) (\Sexpr{release_yr_coef}
kg ha\textsuperscript{-1} year\textsuperscript{-1}). This suggests
that the overall progress on yield over the period from 1980 -- 2016
is approximately \Sexpr{round(release_yr_coef * 36, 0)} kg
ha\textsuperscript{-1}.

When differences between the number of years since each cultivar's
release are taken into account, the estimated increase in grain yield
over time is estimated to be \Sexpr{release_yr_coef}
kg ha\textsuperscript{-1} year\textsuperscript{-1} (90\% CI:
\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]}). Therefore, the evidence is
consistent with no increase in yield potential over the study
period. If the differences in years since release of each cultivar is
ignored, there appears to be a small increase over time (approximately
\Sexpr{round(coef(lm(yield_kg ~ release_yr, mod_data[res,])), 1)})
(Fig. \ref{fig:release}), much less than previous estimates of
\Sexpr{54 * 1.12} kg ha\textsuperscript{-1} year\textsuperscript{-1}
(CITE).  Comparing the two dominate cultivars M-202, which was widely
planted from the late 1980s until the early 2000s, and M-206,
currently the most widely planted CRRFS cultivar (approximately 50\%
of planted area), there is an estimated \Sexpr{m202_v_206} kg
ha\textsuperscript{-1} yield advantage of M-206 over M-202 (Figs
\ref{fig:release} \& \ref{fig:compare_vars}), though both cultivars
yielded similarly at their respective year of release
(Fig. \ref{fig:compare_yrs}).

@@

\subsection{Improving yield potential over time}

Overall, we found evidence consistent with a no improvements in yield
potential in medium grain rice cultivars released by CCRRF's breeding
improvement program over the period \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])}. Rather, it seems possible that the
breeding program has been maintaining yield potential in the face of
changing biotic and abiotic pressures, which are expected to decrease
yield over time \citep{Peng2010}. This highlights both the need
for maintanence breeding and the difficulty of increasing genetic
yield potential as yield approaches the physiological limit. Without
maintanence breeding continuously selecting cultivars well adapted to
current conditions, CA rice production systems are predicted to have
lost an estimated \Sexpr{round(yrs_coef * 31, 0)} kg
ha\textsuperscript{-1} grain yield over the study period.

However, the goal of a breeding program is to not only recover the
benchmark yield of the old standard cultivar but to surpass it. Given
the estimated yield decreases over time, to hit an ambitious target of
500 kg ha\textsuperscript{-1} (approximately 5\%) increase in yield
per release (with roughly 5 years between releases), a new cultivar
needs to yield \Sexpr{round(yrs_coef * -5, -1)} more
in side-by-side comparison on top of the 500 kg ha\textsuperscript{-1}
target improvement. Often side-by-side yield improvements (not
considering the difference in years since release) are closer to 2 --
3\% per release, in which case no absolute improvements might be made
beyond maintenance. The evidence here is consistent with the latter
situation occuring in CA, which suggests that the genetic yield
potential has not increased substantially in the last 30+ years (Fig. \ref{fig:release}).

>>>>

\documentclass[12pt,]{elsarticle}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile, textcomp}
\usepackage{setspace}
\usepackage{etoolbox}
\usepackage{float}
\doublespacing
\usepackage{lineno}
\usepackage{lscape} %% Harvard
\bibliographystyle{model2-names}\biboptions{authoryear}
\pagenumbering{gobble}
%% For figure numbers in supplementary info
%% from http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
\newcommand{\beginsupplement}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
}

\newcommand{\tsups}{\textsuperscript}
\usepackage{Sweave}
\SweaveOpts{prefix.string=graphics/report}

@@

\title{Rice yield improvements by plant breeding are offset by
  decreases over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
\author[ucd]{David MacKill}
\author[ucd]{Jim E. Hill}
\author[res]{Kent McKenzie}
\author[ucd]{Chris van Kessel}
\author[ucd]{Bruce A. Linquist}

@@

Meeting the challenge of feeding an growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite demonstrated increased yield
potential of new genetics over old in side-by-side comparisons. We
analyzed experimental data from one such system, California (CA) rice
production, in order to better understand what improvements have been
made in the genetic yield potential via plant breeding. Specifically,
the hypothesis that a rice variety's yield erodes over time after the
initial selection was tested as an explanation for how side-by-side
comparisons show increased yield while no substantive increases have
been seen in farmers' fields. We confirmed an apparent increase in
yield in side-by-side comparisons between old and newer varieties when
the difference between the years since release was not
considered. However, we demonstrate this apparent increase in yield
can be explained by the differences in the time since selection
between cultivars. The impact of yield erosion was estimated to be
29.8 kg ha\tsups{-1} year\tsups{-1} after initial
selection. Once this effect was considered, the apparent yield
advantage of newer cultivars over old was uncertain. This underscores
the importance of breeding to maintain yield and the challenges of
increasing the genetic yield potential.

@@

\begin{keyword} rice, yield potential, breeding
\end{keyword}

@@

% \linenumbers

@@

load("../data/model_data.rda")
load("h1_fit.Rda")
load("bivar_fit.Rda")
load("../data/all_vt_weather.Rda")

@@

release_yr_coef = round(fixef(RES_full)[2] * scl, 1)
yrs_coef = round(fixef(RES_full)[3] * scl, 1)
only_release_coef = round(fixef(RES_rel_fit)[2]*scl, 1)

RES_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
rel_CI = round(posterior_interval(RES_full, pars = "release_yr_c") * scl, 1)
only_release_CI = round(posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl, 1)

m202_v_206 = round(yrs_coef * -19, -1)

id_est = ranef(RES_full)$id
id_est = cbind(id_est, posterior_interval(RES_full, regex_pars = "id:M"))
idx = order(id_est[,1], decreasing = FALSE)
id_est = id_est[idx,]
@

@@

Constraints on arable land are increasing simultaneous with the need
to increase total production to meet growing demand \citep{Foley2011a,
Godfray2010, Mueller2012}, which has necessitated harvesting more
grain per unit land area \citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple crops,
however more and more production systems are experiencing plateaus in
grain yield \citep{Grassini2013, Peng2010}. If continued improvements
per unit land area are no longer possible, increasing the area under
cultivation will be needed to meet demand, which carries with it
undesirable ecological implications. Therefore, it is of critical
importance to better understand why grain yields have slowed or
plateaued in intensified production systems.

California (CA) rice production represents one such production
system. Rice yields in CA during the period from 1920 -- 1990
increased rapidly (Fig. \ref{fig:state_yield}), in part due to
improvements in rice genetics (i.e., the ``Green Revolution'') through
the efforts of the California Cooperative Rice Research Foundation
(CCRRF; a collaboration between the University of California, the
USDA-Agricultural Research Service, and farmer-funded
research). However, since 1990 CA state average yields have increased
more slowly (Fig. \ref{fig:state_yield}) despite continuous plant
improvement \citep{ccrrf}. For the last 30 years, each newly developed
cultivar has demonstrated a yield advantage over the previous
cultivars in the Statewide Variety Trials (an annual multi-site
variety trial program that compares experimental entries side-by-side
with ``check'' cultivars). Yet, taken across years, these same reports
show stagnant yields. For example, at the year of release (1985),
cultivar M-202 averaged 11 Mg ha\tsups{-1} \citep[14\%
MC;][]{m202}. In 2015, the reported average yield of the newest released
cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep{m209}. Yet this same
report, M-209 demonstrated a X\% yield advantage over M-202. In other rice
systems, a decrease in a cultivar's yield performance over time has
been observed \citep{Peng1998,
Peng2010, Peng2000, Datta1995}, often termed yield erosion
\citep{Peng1998}. It has been speculated that this is due to the
inability of a selected cultivar's genetics to adapt to changing  biotic and
abiotic conditions. We sought to quantify this phenomenon to better
understand how plant improvement and yield erosion are contributing to
the slow yield improvement in farmers' fields over the last
30 years in CA. Specifically, two competing hypotheses were
tested: (1) the yield advantage of each newly developed cultivars is
greater than previous cultivars, and (2) the erosion of yield over
time is creating the illusion of improved yields (i.e., the new
cultivar is recovering the old benchmark as the older cultivar's yield
erodes).

\begin{figure}[H]
<<fig=TRUE, echo=FALSE>>=
CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))/1000
CA_data = CA_data[order(CA_data$Year),]
plot(Value * 1.12 ~ Year, CA_data,
     # type = "l",
     pch = 16,
     xlab = "Year",
     ylab = expression("State avg. grain yield ( Mg ha"^-1~")"))
mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
                  CA_data$Value * 1.12,
                  rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
                rep(1/11,11))
lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)

@
\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
  -- \Sexpr{max(CA_data$Year)} for California, USA, as reported by the
  US Dept. of Agriculture National Agricultural Statistics
  Service. Rapid gains in grain yield were experienced from
  approximately 1950 -- 1990, but annual gains since 1990 have been
  slower. The line represents the 10 year moving average.}
\label{fig:state_yield}
\end{figure}

@@

The CCRRF Rice Experiment Station (RES) is located north-west of
Biggs, CA (39.4648, -121.7342), and has been the central location for
CCRRF's efforts to develop improved rice varieties adapted to CA since
1969. The climate at the RES is Mediterranean, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} -- \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% SOM \citep{ssurgo} in the
top 15cm. As the central location for the CCRRF breeding program, the RES is
researcher-managed (as opposed to farmer-managed trials) and conducts
the bulk of the trials for the Statewide Variety Trials (approximately
3 -- 4 times the number of plots compared to any other location). For
these reasons, we focused our investigation on the RES to avoid
potential complications due to site, climate, and management
differences at the other 6 -- 8 farmer-managed trials per year.

@@

Plots at the RES are managed similar to prescribed best management
practice, including land preparation, fertility management, and pest
and disease control. For planting, rice seed is pre-germinated prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\tsups{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
moisture content prior to reporting.

@@

Yield was modeled as a response to the number of years since the
variety was officially released and the year of release using a
mixed-effects linear regression. To account for similarities between years
and between cultivars, year and cultivar were included in the model as
random effects.

@@

Data was processed and models were fit in R, an environment for
statistical computing \citep{R}. Models were fit using the `rstanarm'
package \citep{rstanarm}, an interface to Stan, a language for
probabilistic programming \citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package \citep{loo},
an efficient means of conducting leave-one-out cross validation
\citep{loo_paper}, with the full model estimated to have slightly
better predictive accuracy compared to the reduced models.

@@

\subsection{Genetic yield improvements}

There was no conclusive evidence that the yields of rice cultivars
selected by CCRRF have, on average, increased over the last 30
years. The estimated change in grain yield per year was estimated to
be a negligible \Sexpr{release_yr_coef} kg ha\tsups{-1}
year\tsups{-1}, with both small positive and negative trends within
the credible interval (90\% CI: \Sexpr{rel_CI[1]} --
\Sexpr{rel_CI[2]}). Hence, the evidence here is consistent with little
to no increase in yield potential over the study period. When yield
erosion is not corrected for in the comparison (i.e., omitted from the
model), there is an apparent small increase
per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} --
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:release}), though this
estimate ignores the confounding effect of yield erosion over
time.

There do appear to be some differences between cultivars, with some
cultivars showing higher yield than the overall average. The marginal
estimate for the performance of a cultivar (averaged over the
uncertainty in the other parameters in the model) shows three
cultivars with higher than average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=', ')}
(Fig. \ref{fig:ids_comp}). A fourth, M-209, might also have higher
than average yield, though the CI overlaps with zero.

\begin{figure}[H]

<<fig=TRUE, echo=FALSE>>=
yrs <- 1975:2025
pp <- posterior_linpred(RES_full, re.form = NA,
                        newdata = data.frame(release_yr_c = yrs - 2000,
                                             yrs_in_trial = 0),
                        prob = 0.5)
pp <-  (pp * scl + center) / 1000
ints = apply(pp, 2, quantile, p = c(0.05, 0.95))

res = mod_data$site == "RES"
comp_yrs = 2010:2015
idx = mod_data$year %in% comp_yrs

plot_data = prep_plot_data(mod_data$yield_kg[idx] / 1000 ,
                           mod_data$id[idx])

plot_data$tmp = tapply(mod_data$release_yr[idx],
                       mod_data$id[idx], unique)

plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = expression("Grain yield ( Mg ha"^-1~")"),
     ylim = c(8, 12.5),
     xlim = c(1980,2019))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * release_yr_coef)) / 1000,
       b = release_yr_coef/1000)
abline(lm(yield_kg/1000 ~ release_yr, data = mod_data[idx,]), lty = 2)
with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.8)

# legend("bottomright", legend = c("w/ years since release adj.",
#                                  "w/o years since release adj.",
#                                  "90% Credible interval"),
#        lty = c(1,2, NA),
#        bty = "n",
#        fill = c(NA,NA, "grey"),
#        border = NA,
#        merge = TRUE,
#        cex = 0.8)

legend("bottomright", legend = c("w/ years since release adj.",
                                 "w/o years since release adj."),
       lty = c(1,2),
       bty = "n",
       border = NA,
       merge = TRUE,
       cex = 0.8)
# polygon(x = c(yrs, rev(yrs)),
        # y = c(ints[1,],rev(ints[2,])),
        # border = NA,
        # col = rgb(0,0,0,0.1))

# text(1990, 8.0, bquote(.(round(release_yr_coef))~kg~ha^-1~year^-1))
@
\caption{The average yield ($\bullet$) and standard error (|) of
medium grain rice cultivars by year of release. For display purposes,
only data for experimental plots planted at the California Cooperative
Rice Research Foundation Rice Experiment Station (California, USA)
over the years \Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)} are
included. Ignoring difference in years since release (dashed line)
suggests an increase in yield potential over time. However, correcting
the comparison to account for yield erosion over time (solid line)
suggests the apparent increase might be due to differences in years
since release.}
\label{fig:release}
\end{figure}

@@

\begin{figure}[H]

@@

plot_data = prep_plot_data(mod_data$yield_kg / 1000,
                           mod_data$yrs_in_trial)

@@

@
\caption{The relationship between yield and the years since release
  for medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. The estimated erosion of yield over time
  (averaged over differences in cultivar and year) is
  estimated to be \Sexpr{yrs_coef} kg ha\tsups{-1}
  year\tsups{-1}. Points ($bullet$) and bars (|) represent the mean and standard
  error of the raw data, respectively. The
  shaded area is the 90\% credible interval.}
\label{fig:RES_year}
\end{figure}

There is estimated to be a small loss of yield over time once a
cultivar is selected (i.e., yield erosion) (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}, 90\% CI:
\Sexpr{RES_CI[2]} -- \Sexpr{RES_CI[1]}). The oldest cultivar in the
trials is estimated to have lost \Sexpr{round(yrs_coef*-36/1000,1)} Mg
ha\tsups{-1} since release. Comparing two dominate
medium grain rice cultivars (M-202, which was widely planted from the late 1980s until
the early 2000s, and M-206, the most widely planted CRRFS
cultivar at approximately 50\% of planted area at the time of
writing), the two cultivars experienced similar trends over time to
each other (Fig. \ref{fig:compare_yrs}) and to the estimate from all
released medium grains (Table \ref{tbl:coef}). There is an estimated
\Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206
over M-202 at the time of writing due solely to the erosion of M-202's
yield over time (Figs \ref{fig:release} \&
\ref{fig:compare_vars}). Further supporting this estimate, both cultivars yielded
similarly at their respective year of release
(Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was released
19 years prior to M-206 (Fig. \ref{fig:release}, Table \ref{tbl:ids}).

\begin{figure}[H]
<<fig=TRUE, width = 8, height = 5, out.width = '85%', echo=FALSE>>=
tars <- c("M202","M206")
par(mfrow = c(1, length(tars)),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

@@

@
\caption{The performance of two major cultivars at the CA Rice
  Experiment Station (RES) over the period
  \Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
  solid line represents a simple linear regression of yield for each
  cultivar over the period which it was planted at the RES (fit to only
  that cultivar's data), while the dashed line is the published 3-year
  average yield of M-202 at its release (1982-1984). Both cultivars have
  a similar yield at their respective year of release, and similar
  decreases over time. The slope for the individual regressions are
  comparable to the slope estimated from all medium grain cultivars
  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}).}
\label{fig:compare_yrs}
\end{figure}

@@

Overall, we do not find conclusive evidence of increased yield
potential increasing in medium grain rice cultivars released by
CCRRF's breeding improvement program over the period
\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}. After correcting yield comparisons
for yield erosion over time in the older cultivar, the estimated yield
improvement spans from \Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]} kg
ha\tsups{-1} year\tsups{-1}.  Even when the
comparison is not corrected for differences in years since release
between cultivars, the estimated improvements over time
(\Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; Fig. \ref{fig:release} dashed line) are
still much less than previous estimates of 37 -- 60 kg
ha\tsups{-1} year\tsups{-1}
\citep{Omar2014}. Given this evidence, we conclude that the breeding
program has likely been maintaining but not substantially improving
yield potential in the face of changing biotic and abiotic
pressures. These biotic and abiotic forces are expected to erode yield
over time \citep{Peng1998, Peng2000, Peng2010}. In fact, the trend
seen here is extremely consistent with previous reports of erosion of
genetic gains of cultivar IR8 at IRRI \citep{Peng1998, Peng2000}. The
evidence here further supports that the erosion of yield over time is
a phenomenon that affects all rice breeding programs. This further
highlights both the need for maintenance breeding and the difficulty
of increasing genetic yield potential as yield approaches the
physiological limit. Without maintenance breeding continuously
selecting cultivars well adapted to current conditions, CA rice
production systems are predicted to have lost an estimated
\Sexpr{round(yrs_coef * (-36/1000), 1)} Mg ha\tsups{-1}
grain yield over the study period.

The goal of a breeding program is to not only maintain the
benchmark yield of the previous ``standard'' but to surpass it. Given
the estimated yield decreases over time, to hit an ambitious target of
0.5 Mg ha\tsups{-1} (approximately 5\%) increase in yield
per release (with roughly 4 -- 5 years between releases), a new cultivar
needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} --
\Sexpr{round(yrs_coef*-5,-1)} kg ha\tsups{-1} in
side-by-side comparison on top of the 0.5 Mg ha\tsups{-1}
target improvement. Often realized side-by-side yield
improvements 
are closer to 2 -- 3\% per release, in which case no absolute
improvements are made beyond maintenance. The evidence here is
consistent with the latter situation occurring in CA, further suggesting
that the genetic yield potential has not increased substantially in
the last 30+ years (Fig. \ref{fig:release}).

\begin{figure}[H]
<<fig = TRUE, echo = FALSE, width=5, height=5>>=

plot(1, type = "n",
     # xaxt = "n",
     xlab = "Years",
     ylab = expression("Grain yield ( Mg ha"^-1~")"),
     ylim = c(8.5,11.5),
     xlim = c(0,40),
     xaxs = "i")
